---
title: "Something New"
author: "Madison Hobbs & Lathan Liou"
class: "MATH150: Methods in Biostatistics"
date: "4/13/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
data <- read_csv("AIDSdata.csv")
```

#Survival Curves
```{r}
fit <- survfit(Surv(time, censor) ~ 1, 
               data = data, 
               type = "kaplan-meier", 
               conf.typ ="log-log", 
               se.fit = TRUE)

#plot KM curve
ggsurvplot(fit, data = data, 
           risk.table = TRUE, 
           conf.int = TRUE, 
           ggtheme = theme_minimal(), 
           risk.table.y.text.col = T, 
           risk.table.y.text = F)
```

#Bootstrapping (Lathan)
##Challenges
Personally, my biggest challenge when learning something new is deciding to what degree I'd like to understand the topic. There is a surface understanding of the definition, a more difficult understanding of the mathematics, and an even more difficult understanding of the conceptual applications. In the case of bootstrap, I think I will find challenging understanding the math behind how bootstrap works.

##Sources
https://www.datacamp.com/community/tutorials/bootstrap-r (Overview of bootstrapping)
Efron, B. (1981) Censored Data and the Boostrap. Journal of the American Statistical Association.
https://stats.stackexchange.com/questions/22017/sample-size-and-cross-validation-methods-for-cox-regression-predictive-models

##A brief overview of Bootstrap and its applications to survival analysis
Bootstrap relies on sampling with replacement of the sample data and is commonly used to estimate the standard error for any statistic. If we had an original sample of $n$ elements,$X$, we resample $X$ $m$ times to get new bootstrap samples ${X_i,...X_m}$ and calculate the desired statistic of each bootstrap sample, ${t(X_i),...t(X_m)}$. 

